Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
broader citation of previous work

The Bradley-Terry model for datasets involving paired comparisons has
wide uptake in the {R} community.  The canonical case is to
consider~$n$ players who compete against one another in pairs; the
basic inference problem is to estimate
numbers~$\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$, $\sum
p_i=1$ so that $\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.
Information about the~$p_i$ may be obtained from the results of paired
comparisons between the players.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features not present in the reified entity
approach of~\citet{hankin2020_jss}.  Firstly, if we consider two players
of identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.

@@

Equation~\ref{mostgeneral} is very general and may be applied to a
range of situations in which standard Bradley-Terry strengths are
modified in some way.  In the following section, I apply it to
Plackett-Luce likelihoods~\citep{plackett1975_jss} likelihood function for
rank data.

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975_jss} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating
conditional independence.  Now we use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple way would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his still-running teammates, boosting their performance.  The
\pkg{hyper2} package includes function \fct{cheering} which
implements this functionality.

>>>>

The Bradley-Terry model for datasets involving paired comparisons has
wide uptake in the {R} community [see, for example,
\citet{turner2012,turner2020} and references therein].  The canonical
case is to consider~$n$ players who compete against one another in
pairs; the basic inference problem is to estimate
numbers~$\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$, $\sum
p_i=1$ so that $\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.
Information about the~$p_i$ may be obtained from the results of paired
comparisons between the players.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and
\pkg{hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this is not
easily represented as a generalized linear model using the techniques
of~\citet{turner2020}.  Relatedlikelihood functions for drawn matches
were considered using the reified entity technique
of~\citet{hankin2020_jss} in which a drawn match was considered to be
won by a fictitious entity (a ``draw monster'') who possessed his own
BT strength.  However, the likelihood function of
table~\ref{windrawloss} has a number of desirable featureas not
present in the reified entity approach of~\citet{hankin2020_jss}.
Firstly, if we consider two players of identical strengths and nullify
the home ground effect (that is, $\lambda=1$) then the probability of
a draw is $\frac{2D}{2+2D}$; the odds of a draw is simply $D$,
regardless of team strength.  Secondly, considering only won (that is,
not score-drawn) games, the probability of a home win is
$\frac{\lambda}{1+\lambda}$, that is, the odds of a home win is simply
$\lambda$, independently of the player strength. Such arguments give
direct operational significance to $\lambda$ and $D$.

@@

Equation~\ref{mostgeneral} is very general and may be applied to a
range of situations in which standard Bradley-Terry strengths are
modified in some way.  In the following section, I apply it to
Plackett-Luce \citep{plackett1975_jss} likelihood functions for
rank data

@@

In the following, I use the ``race'' metaphor: five runners take part
in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
competitor $a$ wins, $b$ comes second, and so on.  Consider the
following Plackett-Luce~\citep{plackett1975_jss} likelihood function for
this observation:
%
\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}
%
Note the use of multiplication between the terms, indicating the
standard assumption~\citep{turner2020} of conditional independence.
Now we use the multiplicative generalization of Bradley-Terry
strengths to introduce an element of non-independence between the
terms.  There are many ways to do this, but one simple way would be to
define equivalence classes of the competitors with each equivalence
class comprising mutually supporting runners.  The metaphor would be
that a runner who has finished the race is able to support other
members of his equivalence class by cheering his still-running
teammates, boosting their performance.  The \pkg{hyper2} package
includes function \fct{cheering} which implements this functionality.
